\section{Future Work and Extensions}
\label{futurework}

The current implementation of SmlCL represents an initial
implementation of a Standard ML interface to OpenCL, that tries to
keep within the functional paradigms of SML, while allowing users to
take advantage of some of the capabilities of modern GPUs. While it
achieves many of its goals by being easy and safe to use, it does not
yet take full advantage of OpenCL and its capabilities, and thus there
is still plenty of room for improvements.

A relative simple addition to the current interface would be to
allow for additional types of buffers and values. Currently, SmlCL
only supports buffers and base values of type \texttt{int} and
\texttt{real} (\texttt{int} and \texttt{double} in C,
respectively). Although C does not have native support for boolean
values like \texttt{true} and \texttt{false}, it does support boolean
logic, simply by regarding the number $0$ as false, and everything
else as true.  Support for SML characters like \texttt{\#"a"} could
also be added rather easily, since for most cases they share the same
representation in Standard ML and C. Strings could also be passed
rather easily, but special care needs to be taken not to allow
any overflows, since different SML implementations have different
representations of strings (they are not null terminated in MLton, for
example). However, adding support for tuples and datatypes is not very
feasible; MLton does not even allow passing datatypes or tuples to C
functions, and it would needlessly complicate kernels as well.

Another trivial extension to SmlCL would be to add more versions of
\texttt{mkKern} and \texttt{kcall}, each allowing a different amount
of buffer arguments. One would need to add run functions to all the
layers of SmlCL, including PrimCL and SimpleCL, but the
implementations would be more or less identical to the already
existing run functions. However, a nicer approach would be leverage
the type-system to allow creations of type-parameterized kernels of
arbitrary arity, and being able to execute them with a single
function. \citet{danvy1998functional} describes a technique for
simulating C's \texttt{printf} function in Standard ML, that could be
used to implement the desired n-arity kernel calls in SmlCL.

One could also allow for static handling of buffer
sizes. By emplying the techniques demonstrated in ~\cite{buffersize}
we can statically guarantee that all kernels are called with suitably
sized buffers.

SmlCL currently only supports passing buffers as arguments to
kernels. A nice addition would be to allow users to pass regular
scalars like integers or reals. However, this poses several problems
for the implementation: firstly, we would need to extend the function
definitions of \texttt{kcall1} and \texttt{kcall2} to allow not only
buffers, but also integers and reals as arguments. Sadly, this is not
easily possible using the current way we execute functions: we would
need to add some datatypes to manage the different types of arguments,
and we would have to add a lot of foreign function calls for each
composition of argument types. Another way to achieve this would be to
refactor the code, such that we had functions like:

\begin{lstlisting}[language=ML,mathescape]
  val setArgBuf : kernel -> int -> 'a buf -> unit;
  val setArgReal : kernel -> int -> real -> unit;
  val setArgInt : kernel -> int -> int -> unit;
\end{lstlisting}

That is, we manually set the kernel arguments on the SML side, instead
of letting the C-side handle them.

We could also extend our interface to OpenCL, allowing users to
specify work-group sizes, adding support for OpenCL vectors, using
kernels that have been compiled offline, use CPUs in addition to GPUs,
and much more. OpenCL is very complex, and has many options for
customization of execution and how to run your programs. However, we
feel that many of these additions would work counter to our goals of a
simplified API that hides some of this complexity in order to provide
ease-of-use. It would however enable users to better take advantage of
the capabilities of OpenCL, and if a suitable compromise could be
found, it could facilitate more efficient parallel processing.

Finally, we could extend the expression generator, in order to allow
the construction of more complex kernels, including additional
constructions like \texttt{if}-statements, \texttt{for} and
\texttt{while} loops, auxiliary functions, global constants, local and
private variables, and so forth. This could be done in a number of
ways, one could either provide sufficient constructs such that our
expression interface mimics OpenCL C, or one could keep a functional
approach and add more constructs like \texttt{Map} and \texttt{Fold},
\texttt{Let} statements, and the like to the language. For this
project, we have chosen to keep it relatively simple, in order to make
sure that kernels execute correctly. The expression generator could
also be extended to allow for kernel composition, such that one could
be combined two or more kernels into one execution, effectively
creating a kernel algebra. In that way you could combine, say a map
operation and a reduce operation into a single operation, saving some
transfer-time between the device and the host. Such an algebra would
also give incentive to create more kernel templates like redce. An
obvious addition to the current kernel templates, could be a kernel
that creates an of values directly on the device.
